---
title: "R Data Visualization Basics"
author: "Ramberto Jr. Sosa Cueto"
date: "`r Sys.Date()`"
output: html_document
---

## Aesthetics for points

# ggpplot Packages and dataset 

```{r Install Packages + Palmer Penguin dataset}
install.packages("ggplot2")
install.packages("palmerpenguins")
```

```{r Loading Packages + Palmer Penguins dataset}
library("ggplot2")
library("palmerpenguins")
```

# ggplot visualization - plus (+) signs is use to add layers to the plot; geom_point to create points to represent data; and mapping=aes refers to y and x axis, how the data will look. Mapping of the data

```{r flipper length vs body mass plot}
ggplot(data=penguins) + geom_point(mapping = aes(x=flipper_length_mm, y=body_mass_g))
```

# adding "colors" to the aesthetic (aes) to identify "species" in palmer penguins

```{r flipper length vs body mass plot}
ggplot(data=penguins) + geom_point(mapping = aes(x=flipper_length_mm, y=body_mass_g, color=species))
```

# adding "shapes" to the aesthetic (aes) to identify "species" in palmer penguins

```{r flipper length vs body mass plot}
ggplot(data=penguins) + geom_point(mapping = aes(x=flipper_length_mm, y=body_mass_g, shape=species))
```

# adding "colors" and "shapes" to the aesthetic (aes) to identify "species" in palmer penguins

```{r flipper length vs body mass plot}
ggplot(data=penguins) + geom_point(mapping = aes(x=flipper_length_mm, y=body_mass_g, shape=species, color=species))
```

# adding "size", "colors" and "shapes" to the aesthetic (aes) to identify "species" in palmer penguins

```{r flipper length vs body mass plot}
ggplot(data=penguins) + geom_point(mapping = aes(x=flipper_length_mm, y=body_mass_g, shape=species, color=species, size=species))
```

# using alpha with "species" to control transparency of data

```{r flipper length vs body mass plot}
ggplot(data=penguins) + geom_point(mapping = aes(x=flipper_length_mm, y=body_mass_g, alpha=species))
```

# changing color for all data points. 

```{r flipper length vs body mass plot}
ggplot(data=penguins) + geom_point(mapping = aes(x=flipper_length_mm, y=body_mass_g), color="purple")
```

## Geom Functions, point, bar, line etc...

# geom_smooth

```{r flipper length vs body mass plot}
ggplot(data=penguins) + geom_smooth(mapping = aes(x=flipper_length_mm, y=body_mass_g))
```

# adding additional geoms 

```{r flipper length vs body mass plot}
ggplot(data=penguins) + geom_smooth(mapping = aes(x=flipper_length_mm, y=body_mass_g)) + geom_point(mapping = aes(x=flipper_length_mm, y=body_mass_g)) 
```

# adding line type to identify different species

```{r flipper length vs body mass plot}
ggplot(data=penguins) + geom_smooth(mapping = aes(x=flipper_length_mm, y=body_mass_g, linetype=species))
```

# using jitter function, to makes points easier to find when data overlaps

```{r flipper length vs body mass plot}
ggplot(data=penguins) + geom_jitter(mapping = aes(x=flipper_length_mm, y=body_mass_g))
```

# Bar Chart with Diamon data, When "y" axis isn't specify it automatically turns into COUNT (or Data Frenquency)

```{r Plot Diamond Cut}
ggplot(data=diamonds) + 
  geom_bar(mapping = aes(x=cut))
```

# adding color to the cut

```{r Plot Diamond Cut}
ggplot(data=diamonds) + 
  geom_bar(mapping = aes(x=cut, color=cut))
```

# using fill to add color inside bar for cut

```{r Plot Diamond Cut}
ggplot(data=diamonds) + 
  geom_bar(mapping = aes(x=cut, fill=cut))
```

# using fill to add color inside bar for clarity. The result is a stacked bar identifying clarity inside cut bars to identify volume

```{r Plot Diamond Cut}
ggplot(data=diamonds) + 
  geom_bar(mapping = aes(x=cut, fill=clarity))
```

## Using facets, to display smaller groups or subsets in data

# facet_wrap function, using species

```{r flipper length vs body mass plot}
ggplot(data=penguins) + geom_point(mapping = aes(x=flipper_length_mm, y=body_mass_g, color=species)) + facet_wrap(~species)
```

# facet_wrap function, using cut from Diamond dataset

```{r Plot Diamonds}
ggplot(data=diamonds) + geom_bar(mapping=aes(x=color, fill=cut)) + facet_wrap(~cut)
```

# facet_grid function, using with penguins dataset. Must usefull when exploring relationship between multiple groups; sex and species

```{r flipper length vs body mass plot}
ggplot(data=penguins) + geom_point(mapping = aes(x=flipper_length_mm, y=body_mass_g, color=species)) + facet_grid(sex~species)
```

# using facet_grid to look at different species

```{r flipper length vs body mass plot}
ggplot(data=penguins) + geom_point(mapping = aes(x=flipper_length_mm, y=body_mass_g, color=species)) + facet_grid(~species)
```

# using facet_grid to look for sex

```{r flipper length vs body mass plot}
ggplot(data=penguins) + geom_point(mapping = aes(x=flipper_length_mm, y=body_mass_g, color=species)) + facet_grid(~sex)
```

## Labels & Annotations

# adding title using labs

```{r flipper length vs body mass plot}
ggplot(data = penguins) + 
  geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  labs(title = "Palmer Penguins: Body Mass vs. Flipper Length")
```

# adding subtitle and caption

```{r flipper length vs body mass plot}
ggplot(data = penguins) + 
  geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  labs(title = "Palmer Penguins: Body Mass vs. Flipper Length", subtitle = "Sample of Three Penguin Species", caption="Data collected by Dr. Kristen Gorman")
```

# using annotate function, the location is specify using x and y axis values

```{r flipper length vs body mass plot}
ggplot(data = penguins) + 
  geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  labs(title = "Palmer Penguins: Body Mass vs. Flipper Length", subtitle = "Sample of Three Penguin Species", caption="Data collected by Dr. Kristen Gorman") +
  annotate("text", x=220, y=3500, label="The Gentoos are the largest")
```

# adding color, size and angle to annotations

```{r flipper length vs body mass plot}
ggplot(data = penguins) + 
  geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  labs(title = "Palmer Penguins: Body Mass vs. Flipper Length", subtitle = "Sample of Three Penguin Species", caption="Data collected by Dr. Kristen Gorman") +
  annotate("text", x=220, y=3500, label="The Gentoos are the largest", color="purple", fontface="bold", size=4.5, angle=0)
```

# saving plot. Usefull for adding layers. Example with annotations

```{r saving chart in a variable}
p <- ggplot(data = penguins) + 
  geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  labs(title = "Palmer Penguins: Body Mass vs. Flipper Length", subtitle = "Sample of Three Penguin Species", caption="Data collected by Dr. Kristen Gorman")

p + annotate("text", x=220, y=3500, label="The Gentoos are the largest", color="purple", fontface="bold", size=4.5, angle=0)
```

## Saving Visualizations

```{r flipper length vs body mass plot}
ggplot(data=penguins) + geom_point(mapping = aes(x=flipper_length_mm, y=body_mass_g, shape=species, color=species))
```

# using ggsave the last visual is save

```{r}
ggsave("Three Penguin Species.png")
```




